---
title: "Shiny 101: The Modular App Blueprint"
author: "Veerle van Leemput"
format:
  revealjs: 
    theme: [default, custom.scss]
    footer: "[github.com/hypebright/shinyconf2024-shiny101](https://github.com/hypebright/shinyconf2024-shiny101)"
    logo: images/logo.png
include-in-header: 
  text: |
    <link href="https://cdn.jsdelivr.net/npm/bootstrap@5.0.2/dist/css/bootstrap.min.css" rel="stylesheet" integrity="sha384-EVSTQN3/azprG1Anm3QDgpJLIm9Nao0Yz1ztcQTwFspd3yD65VohhpuuCOmLASjC" crossorigin="anonymous">
highlight-style: "nord"
---

## Program

- **19:30 - 20:00** Introduction to Shiny
- **20:00 - 20:30** Hands-on: Building a Shiny App Together
- **20:30 - 20:40** Break ☕️
- **20:40 - 21:00** Introduction to Shiny modules
- **21:00 - 21:30** Hands-on: Implementing Shiny modules
- **21:30 - 21:45** Sharing and feedback
- **21:45 - 22:00** Recap and Questions

## Learning objectives

After this workshop, you will know:

- the basics of a Shiny application
- how to make your app modular
- how to communicate between modules

You will leave this workshop with:

- a blueprint for a modular Shiny app

# Introduction to Shiny

## Shiny app basics

Two parts of a Shiny app:

- **UI**: User Interface
- **Server**: Server logic

They come together in a single script: `app.R`

## The app.R script

```{.r code-line-numbers="1"}
library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

## The app.R script

```{.r code-line-numbers="3-5"}
library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

## The app.R script

```{.r code-line-numbers="7-8"}
library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

## The app.R script

```{.r code-line-numbers="10"}
library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

## The ui part

```{.r code-line-numbers="3-5"}
library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

`fluidPage()` creates a Bootstrap page

## The ui part

- Bootstrap is a free front-end framework used in web development and it includes all the basic to build an app that will work in a web browser on any device. 
- It is built on **HTML**, **CSS** and **JavaScript**
- There are many version of Bootstrap, the latest being version 5, but Shiny uses version 3


## The ui part

Under the hood, every UI function is translated to HTML:

```{r}
library(shiny)
```

```{r}
#| echo: true
as.character(titlePanel("Hello, Shiny!"))
```

<br>

This HTML can have styles attached to it, either by the `style` attribute, or by a CSS class.

```{r}
#| echo: true
as.character(sidebarPanel())
```


## The ui part

To let users interact with your app, you can add **input controls**. A basic input control has:

- an `inputId`
- a `label`
- a `value`

## The ui part

For example:

```{.r}
numericInput(inputId = "number", 
             label = "Enter a number", 
             value = 0)
```
<br>
```{.r}
actionButton(inputId = "button", 
             label = "Click me")
```

## The server part

When the app runs, every user gets served the same HTML from the ui.

When the user interacts with the UI, we want that HTML to react to the user.

The server logic uses **reactive programming**: when an input changes, all related outputs react and are updated automatically.

That is possible because these outputs are  **reactive outputs**

## The server part

Shiny has several functions that turn R objects into reactive outputs for your ui: the *Output family

Each function creates a specific type of output, for example:

| Function | Output type |
|----------|-------------|
| textOutput() | text |
| tableOutput() | table |
| plotOutput() | plot |
| uiOutput() | raw HTML |
: {.table .table-striped .table-hover}

## The server part

Every output element needs a single argument: `outputId`- which is a simple string that needs to be **unique**.

```{.r}
textOutput(outputId = "text")
```

This `textOutput `tells Shiny where to display the object

<br>

Next step: tell Shiny how to build the object

## The server part

The server function builds an object named **output** and this object will contain all the code needed to display the R objects in your app

This output object is **list-like**

Each R object that you want to display has its own entry in this list, therefore the name of every output in your ui needs to match a definition in the server

```{.r}
output$text
```

## The server part

Each entry in the output list should contain a render* function. You must use render* functions that match the *Output functions:

| Function | Output type | Render function |
|----------|-------------|-----------------|
| textOutput() | text | renderText() |
| tableOutput() | table | renderTable() |
| plotOutput() | plot | renderPlot() |
| uiOutput() | raw HTML | renderUI() |
: {.table .table-striped .table-hover}

## The server part

```{.r code-line-numbers="2-3"}
server <- function(input, output, session) {
  output$text <- renderText({
  })
}
```

## Using inputs

This works the same with inputs!

<br>

The server can access an object named `input` that will contain the value of all the objects defined in the ui

This object is **list-like**, just like the output object. Each input value has its own entry in this list:

```{.r}
input$number
```

## Connecting the dots

If we combine the input and output objects, we can create a simple app that displays the square of a number 👏.

```{.r}
library(shiny)

ui <- fluidPage(
  numericInput(inputId = "number", 
               label = "Enter a number", 
               value = 0),
               
  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  })
}

shinyApp(ui, server)
```

## Connecting the dots

What about an `actionButton`? We could use this button as an **event** that triggers the calculation.

```{.r}
library(shiny)

ui <- fluidPage(
  numericInput(inputId = "number", 
               label = "Enter a number", 
               value = 0),
               
  actionButton(inputId = "button", 
               label = "Click me"),
               
  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  }) |> bindEvent(input$button)
}

shinyApp(ui, server)

```

# It's all about the looks 💅

Only our app doesn't have any yet... 😢

## Quickly leveling up your UI

- There are many ways to make your app look instantly better
- One of them is using `bslib` which is a package that provides a modern UI toolkit for Shiny and R Markdown based on Bootstrap
- The cool thing: you can get around Shiny's default usage of Bootstrap version 3!

## Working with {bslib}

```{.r}
library(shiny)
library(bslib)

ui <- page_navbar(
  theme = bs_theme(version = 5),
  title = "Modular App Blueprint",
  nav_panel(
    title = "Numbers",
    numericInput(inputId = "number",
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = "button",
                 label = "Click me"),
    textOutput(outputId = "text")
  )
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  }) |> bindEvent(input$button)
}

shinyApp(ui, server)
```

## Working with {bslib}

```{.r code-line-numbers="4-14,17"}
library(shiny)
library(bslib)

custom_theme <- bs_theme(
  version = 5,
  bg = "#F9F9F9",
  fg = "#003f5c",
  primary = "#bc5090",
  secondary = "#58508d",
  warning = "#ffa600",
  danger = "#ff6361",
  info = "#0091d5",
  base_font = font_google("PT Sans")
)

ui <- page_navbar(
  theme = custom_theme,
  title = "Modular App Blueprint",
  nav_panel(
    title = "Numbers",
    numericInput(inputId = "number",
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = "button",
                 label = "Click me",
                 width = "100px"),
    textOutput(outputId = "text")
  )
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  }) |> bindEvent(input$button)
}

shinyApp(ui, server)
```

## Shiny app recap

We covered:

- the ui part
- the server part
- how to connect the two
- how to bind an event to an output

What's next?

- how to use data
- reactives & reactive values

# Hands-on: building a Shiny app together 💻

# Shiny modules 

## What are modules?

You could call modules **special functions**.

<br>

Using functions in a Shiny app is common:

- You use functions in your UI (e.g. `value_box()`)
- You use functions in your server

<br>

That works well for code that is **completely on the ui or server side**.

## What are modules?

For code that spans both the ui and the server, you need a new technique: **modules**

<br>

A module consists of some **UI code** that works together with some **server code**

<br>

Modules are special because they create their own **namespace**: things like input and output ids are isolated from the rest of the app

## Why use modules?

There are two main reasons to work with modules:

1. Because a module creates its own **namespace**, you can write and run code in isolation. You don’t need to worry about what’s going on outside the module.

2. Because a module is basically a **function**, it comes with all the benefits that functions have.

## What does a module look like?

A module has two parts, and is basically a mini-app:

1. **The module UI**, which generates the UI

2. **The module server**, which runs code inside the `server()` function

## What does a module look like?

```{.r}
library(shiny)

ui <- fluidPage(
  numericInput(inputId = "number",
               label = "Enter a number",
               value = 0),

  actionButton(inputId = "button",
               label = "Click me"),

  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  }) |> bindEvent(input$button)
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## What does a module look like?

```{.r}
library(shiny)

numberModUI <- function(id) {
  ns <- NS(id)
  tagList(
    numericInput(inputId = ns("number"),
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = ns("button"),
                 label = "Click me"),
    textOutput(outputId = ns("text"))
  )
}

numberModServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$text <- renderText({
      input$number^2
    }) |> bindEvent(input$button)
  })
}

ui <- fluidPage(
  numberModUI("numbers")
)

server <- function(input, output, session) {
  numberModServer("numbers")
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## What does a module look like?

```{.r code-line-numbers="3-13"}
library(shiny)

numberModUI <- function(id) {
  ns <- NS(id)
  tagList(
    numericInput(inputId = ns("number"),
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = ns("button"),
                 label = "Click me"),
    textOutput(outputId = ns("text"))
  )
}

numberModServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$text <- renderText({
      input$number^2
    }) |> bindEvent(input$button)
  })
}

ui <- fluidPage(
  numberModUI("numbers")
)

server <- function(input, output, session) {
  numberModServer("numbers")
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## What does a module look like?

```{.r code-line-numbers="4,11"}
library(shiny)

numberModUI <- function(id) {
  ns <- NS(id)
  tagList(
    numericInput(inputId = ns("number"),
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = ns("button"),
                 label = "Click me"),
    textOutput(outputId = ns("text"))
  )
}

numberModServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$text <- renderText({
      input$number^2
    }) |> bindEvent(input$button)
  })
}

ui <- fluidPage(
  numberModUI("numbers")
)

server <- function(input, output, session) {
  numberModServer("numbers")
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## What does a module look like?

```{.r code-line-numbers="15-21"}
library(shiny)

numberModUI <- function(id) {
  ns <- NS(id)
  tagList(
    numericInput(inputId = ns("number"),
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = ns("button"),
                 label = "Click me"),
    textOutput(outputId = ns("text"))
  )
}

numberModServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$text <- renderText({
      input$number^2
    }) |> bindEvent(input$button)
  })
}

ui <- fluidPage(
  numberModUI("numbers")
)

server <- function(input, output, session) {
  numberModServer("numbers")
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## What does a module look like?

```{.r code-line-numbers="23-29"}
library(shiny)

numberModUI <- function(id) {
  ns <- NS(id)
  tagList(
    numericInput(inputId = ns("number"),
                 label = "Enter a number",
                 value = 0),
    actionButton(inputId = ns("button"),
                 label = "Click me"),
    textOutput(outputId = ns("text"))
  )
}

numberModServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$text <- renderText({
      input$number^2
    }) |> bindEvent(input$button)
  })
}

ui <- fluidPage(
  numberModUI("numbers")
)

server <- function(input, output, session) {
  numberModServer("numbers")
}

shinyApp(ui, server)

```

:::footer
[/templates/03_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/03_modules)
:::

## Where do you put modules?

You can simple put the module functions in the `app.R` file (you call this inline).

<br>

Other options:

 - Put the module in a separate R script in the /R folder. **Shiny will automatically source it**.

- Put the module in a separate R script in any other folder, and source it using `source(“./my_modules/random.R”)`. 

## Nested modules

Modules are **composable**, and it may make sense to create a module that itself contains a module: this is called **nesting**

<br>

This sounds complicated, but isn’t any different compared to calling a function within another function

## Nested modules

test

:::footer
[/templates/04_nested_modules](https://github.com/hypebright/shinyconf2024-shiny101/blob/main/templates/04_nested_modules)
:::


